A Stopped Negative Binomial Distribution

### 2016-06-22 stat-and-prob-letters/snb-paper.tex
Another day another version.

Consider a prototypical early phase, single-arm clinical trial in which 
12 patients
are enrolled and treated. The binomial probability of success is $p=0.045$ for
each patient under the null hypothesis that treatment is not effective.
If two or more patients out of these 12 respond to the treatment then we 
reject this null
hypothesis and the treatment is deemed successful at significance 
level of $0.1$.  If fewer than two respond then the null is not rejected
and the treatment is judged as ineffective.

If all 12 patients are enrolled at once, as in the classic
design, then the sample size is 12. However, in most clinical trials the
patients are enrolled sequentially, often with one patient's outcome realized
before the next one enters the trial. In the present example, observing two
successful patients allows us to reach one endpoint so the sample required
could be as small as two. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as a predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling the range of the
sample size for this trial is between two and 12.

Let us assume each patient outcome can be modeled as an independent,
identically distributed Bernoulli($p$) random variable. The trial is realized
as a sequence of these random variables stopping when either a
specified number of successes or failures has been reached.

@@

A hypothetical sample path is illustrated in Fig.~\ref{fig:kane_viz}.
The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients 
enrolled. The horizontal and vertical boundaries represent
endpoints for the trial. In this case two successes were reached after
enrolling 10 patients: one in the fourth outcome and on in the eleventh.
Since the success boundary is reached, we say that the trial succeeds,
or more correctly, the treatment succeeds.

@@

\begin{figure}[bp!]
\centering
\subfloat[enrollment distribution]{\includegraphics[width=0.5\textwidth]{snb-first-plot.pdf}}
\hfill
\subfloat[effect of varying $p$]{\includegraphics[width=0.5\textwidth]{mean-and-variance.pdf}}
\caption{
The distribution of a trial that stops after seven patients respond
or 11 patients do not. (a) shows the distribution when
$p = 0.2$. (b) shows the result of varying $p$ between 
zero and one.
}
\label{fig:exp-and-var}
\end{figure}

More generally the distribution for number of trial enrollments is shown in 
Fig.~\ref{fig:first-plot} (a). The probability mass is concentrated at the 
$11^{\text{th}}$ and $12^{\text{th}}$ step since $p$ 
is small and for most realizations reach 11 
non-responders before reaching 2 responders.

Fig.~\ref{fig:exp-and-var} shows the expected value and variance for the
number of trial enrollments varying $p$ between zero an 1. When $p$ is
small the probability of an individual response is small, the treatment
is more likely to fail, and the number of enrollees is large. When $p$
is large the treatment is more likely to succeed and the number of 
enrolees is small. Since $p=1$ and $p=0$ represent the deterministic
the corresponding variance is zero. The variance is right-skewed because
the support of the distribution corresponding to successful treatment
is larger than that of failure.

@@

In the rest of this work we derive the distribution of the number of 
enrollees needed
to observe either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB). Some of its characteristics are
summarized in Tab.~\ref{tab:snb}.
This paper derives these results
and explores properties of the distribution.
%The next section introduces our notation and basic results
%including the density of the distribution along with a description of
%its relation to other distributions. 
Section 2 derives the distribution function
based on a defined Bernoulli process and gives some basic properties.
Section 3 shows how the distribution is related to other standard
distributions and connects the SNB tail probability to the Binomial tail 
probability.
Section 4 derives the moment generating function.
Section 5 derives the compound distribution using a Beta distribution
for $p$.

@@

Let $\,b_1, b_2, \ldots \,$ denote a sequence of independent, identically
distributed, Bernoulli random variables with $\mathbb{P}[b_i=1]=p$ and
$\mathbb{P}[b_i = 0] = 1-p$, for
probability parameter $0\leq p \leq 1$. In the clinical trial setting
$\,b_i = 1$ corresponds to a successful patient outcome following treatment.  
Let $s$ and $t$ be positive integers.  Define the SNB random
variable $Y$ as the smallest
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either}
$\,s\,$ successes {\em or} $\,t\,$ failures. That is, the SNB distribution
of $Y$ is the smallest integer such that either
$\sum_i^Y b_i = s$ or $\sum_i^Y 1-b_i = t$.

The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}.
\end{equation*}
The probability mass function is
\begin{equation} \label{eqn:pmf}
\mathbb{P} [Y=k] = S(k, p, s) \ I_{\{s \leq k\}} + 
  S(k, 1-p, t) \ I_{\{ t \leq k \}}
\end{equation}
where $I_{\{f\}}$ is the {\emph indicator function}, taking the value 
of one if $f$ is true and zero otherwise, and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is the negative binomial probability mass.

To prove (\ref{eqn:pmf}), consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
At each step a patient's outcome is measured. In Fig.~\ref{fig:kane_viz} 
we consider a graphical illustration of the plot $X_k$ against
$k$. If the outcome of the $k$th patient is a success then the process 
advances diagonally in the positive horizontal and vertical direction. 
If the $k$th patient fails
then the sample path advances in the positive horizontal direction only. The
process continues until either $X_k = s$ or $X_k = k-t$.

@@

The probability a given realization of $\mathbf{X}$ reaches $s$ at
the $k$th outcome is the probability that, at time $k-1$, there are $s-1$
successful outcomes and $k-s$ unsuccessful outcomes multiplied by
the probability of the final success at time $k$. This expression is given
in (\ref{eqn:N}). 
Similarly, the probability a given realization reaches $k-t$
is the probability that, at outcome $k-1$, there are $k-t$ successful outcomes
and $t-1$ unsuccessful outcomes multiplied by the probability of a final
unsuccessful outcome at time $k$.

@@

Next, we consider an interim analysis of a clinical trial after $s'$ 
successes and $t'$ failures have been observed for $s' < s$ and $t' < t$.
\begin{corollary} \label{conditional_distribution}
The number of subsequent enrollments needed 
to reach either endpoint behaves as SNB($p$, $s-s'$, $t-t'$).
\end{corollary}
Having observed $s'$
successes and $t'$ failures, there are $s-s'$ successes needed to reach the
success endpoint and $t-t'$ failures needed to reach the failure endpoint.

\section{Connections and Approximations by Other Distributions}

\begin{figure}[p!]
\begin{center}
\includegraphics[width=\textwidth]{shapes.pdf}
\end{center}
\caption{Different shapes of the SNB distribution with parameters 
($p$, $s$, $t$), as given. Black indicates mass contributed by reaching
$s$ successes before $t$ failures. Grey indicates
mass contributed by reaching $t$ failures first. \label{shapes.fig}}
\end{figure}

The SNB is a generalization of the negative 
binomial distribution. If $t$ is large then $Y-s$ has a 
negative binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j]        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is small. As a result, with proper parameter choice, the SNB
can mimic other probability distributions in a manner similar to 
those described in \cite{Peizer1968} and \cite{Best1974}. Examples are
shown in in Fig.~\ref{shapes.fig}.

The SNB also generalizes both the minimum (riff-shuffle) and maximum negative
binomial distributions up to a translation of the support.
For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1970}.
Similar derivations of the closely-related maximum negative binomial 
discrete distributions also appear in~\cite{Zhang2000}
and \cite{Zelterman2005}.
The maximum negative binomial is the smallest number of outcomes necessary to 
observe at least $s$ successes {\em and} $s$ failures. The SNB is the 
number of outcomes to observe {\em either} $s$ successes {\em or} $t$ failures.

@@

There is a close connection between the tail probabilities of the SNB and the 
binomial distributions.
The probability of reaching the success endpoint in an 
SNB($p$, $s$, $t$) random variable is 
equal to the probability of at least $s$ successes in a Binomial distributed 
random variable with size $s+t-1$ and success probability $p$. 
Likewise, the probability of reaching the failure endpoint is equal
to the probability of at most $s-1$ successes in a Binomial distribution. 
That is, the probability that the number of successes is at least $s$ in the 
Binomial model is the same that the treatment succeeds (reaches $s$) in the SNB 
model.
\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed Binomial with size $n=s+t-1$ and success probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [\text{treatment succeeds}].
\end{equation}
\end{prop}
\begin{proof}
The Binomial tail probability is
\begin{equation*}
\mathbb{P}[B \geq s] = 1 - \mathcal{I}_{1-p}.
\end{equation*}
%\begin{align*}
%\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \mathcal{I}_{1-p}.
%\end{align*}
The corresponding SNB probability is
\begin{equation*}
\mathbb{P} [\text{treatment succeeds}] 
  = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{equation*}
Let $i=k-s$. Using the fact that
\begin{equation*}
{i+s-1 \choose s-1} = {i+s-1 \choose i}
\end{equation*}
the last summation can be rewritten as
\begin{align}
\mathbb{P} [\text{treatment succeeds}] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align}
completing the proof.
\end{proof}

To illustrate this result, let us return to our initial example
where $s=2$, $t=11$, and $p=0.045$.  The probability masses in
Fig.~\ref{fig:snb_bin_compare} represented in 
black are equal in panels a) and b) as are the masses in grey.
The probability that $s$
successes are reached in the SNB process is the same as the binomial 
probability of at least two successes. Likewise, the probability that $t$ 
failures are reached in the SNB process is the same as the binomial
probability of zero or one successes.

\begin{figure}[t!]
\centering
\subfloat[SNB distribution]{\includegraphics[width=0.5\textwidth]{snb_density.pdf}}
\hfill
\subfloat[Binomial distribution.]{\includegraphics[width=0.5\textwidth]{bin_density.pdf}}
\caption{
SNB(0.045, 2, 11) with mass contributed from 
$s$ successes (black) or $t$ failures (grey) along with Bin(0.045, 12) with
at least 2 successes (black) or fewer (grey).
}
\label{fig:snb_bin_compare}
\end{figure}

@@

The SNB's ability to approximate the geometric, normal, gamma, and poisson
distributions follow from it generalizing the negative binomial. To 
recover the MGF of the negative binomial consider the case where
$t \rightarrow \infty$ in (\ref{eqn:mgf}). The regularized incomplete
beta function in the first term goes to one and zero in the second term.
We are left with the MGF of the negative binomial distribution.

When $s=1$ this is the geometric distribution. The negative
binomial can therefore be seen as a sum of i.i.d. geometric distributions.
For an appropriately large number of samples the central limit theorem
yields a normal approximation.

Drawing connections to the gamma and poisson distributions are more complicated.
However a connection to the gamma distribution well-studied problem in the 
literature (see \cite{Ord1968,Guenther1972,Best1974} for examples). 
A connection to the poisson appears in \cite{Anscombe1950} where
it is shown that if the mean of a poisson is proportional
to a $\chi^2_{2k}$ distribution then the negative binomial is obtained. 
All of these results work by equating cumulants and then showing 
that differences between betwen the cumulant generating
functions converge to zero.

@@

Let us consider a Bayesian analysis with a Beta distribution on $p$.
\begin{prop} \label{prop:bayesian}
The compound distribution of the Stopped Negative Binomial 
distribution when $p$ is distributed Beta($\alpha$, $\beta$) is
\begin{align} \label{eqn:posterior}
\mathbb{P} [Y = k | s, t, \alpha, \beta ] &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose t-1} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}.
\end{align}
\end{prop}
\begin{proof}
For notational simplicity, assume that $k$ is in the range
$\min(s,t) \leq k \leq s+t-1$. When 
this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose t-1} p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose t-1} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp.
\end{align*}
The result in (\ref{eqn:posterior}) follows by the definition of the 
Beta function.
\end{proof}

To better understand the compound distribution, let us once again return to 
our example where $s=2$, $t=11$. However, we assume $p$ is a Beta distribution
with shape parameters $\alpha = 0.045 c$ and 
$\beta = 0.955 c$ for any $c > 0$.
This parameterization allows us to examine the relationship between 
uncertainty in $p$ and the SNB.

\begin{figure}[h!]
\centering
\subfloat[mean]{\includegraphics[width=0.5\textwidth]{bayesian-sample-expectation.pdf}}
\hfill
\subfloat[variance]{\includegraphics[width=0.5\textwidth]{bayesian-sample-variance.pdf}}
\caption{
The expected size and variance of the compound SNB with $s=2$, $t=11$
with $\mathbb{E}\ p = 0.045$ varying the uncertainty 
in the distribution of $p$. A larger value of $c$
corresponds to a Beta with smaller variance.
}
\label{fig:bayesian-sample-size}
\end{figure}

Fig.~\ref{fig:bayesian-sample-size} shows how the expected sample size
and variance of the compound SNB. As the certainty increases the 
compound distribution sample values converge to corresponding expected 
values with 
$p = 0.045$. It may be noted that the compound distribution's mean
approaches the deterministic limit slowly. This
is because when $c$ is small, the distribution of $p$ is heavily skewed. 
As $c$ increases this effect is minimized.

Slow convergence in the mean will occur when $p$ takes values close to 
zero or one. In these cases it may be preferable to select shape parameters
based on the median or mode of the beta distribution. These values
will vary less in the uncertainty of the distribution.

@@

%A second avenue for methodological innovation comes directly from
%Corollary \ref{conditional_distribution} and Proposition \ref{prop:bayesian}.
%Our knowledge of the response probability $p$ comes from published reports
%yielding a prior distribution.
%While a trial is being conducted the response probabilities are 
%updated as new results are received. The number of individual successes
%and failure are used to learn the shape characteristics of the corresponding
%Beta distribution and resulting posterior
%distribution. An estimate of the probability of reaching the success (and
%failure) endpoints can be provided at each step of the trial.

>>>>

Consider a prototypical early phase, single-arm clinical trial in which 
17 patients
are enrolled and treated. The binomial probability a patient 
responds to treatment is $p=0.2$ 
under the null hypothesis that the treatment is not effective.
If seven or more patients out of these 17 respond to the treatment then we 
reject the hypothesis and the treatment is deemed successful at 
significance level of $0.1$.  If fewer than seven respond then the null is not 
rejected and the treatment is deemed ineffective.

If all 17 patients are enrolled at once, as in the classic
design, then the sample size is 17. However, in most clinical trials the
patients are enrolled sequentially with one patient's outcome realized
before the next one enters the trial. In the present example, observing seven
successful patients is one endpoint and the number of enrollees required
could be as small as seven. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as any predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling the range of the
sample size for this trial is between seven and 17.

Let us assume each patient outcome can be modeled as an independent,
identically distributed Bernoulli($p$) random variable. The trial is realized
as a sequence of Bernoulli samples that stops when either a
specified number of responders or non-responders have been reached.

@@

A hypothetical sample path is illustrated in Fig.~\ref{fig:kane_viz}.
The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients 
enrolled. The horizontal and vertical boundaries represent
endpoints for the trial. In this case a seventh response was reached on
the $15^{\text{th}}$ enrollment.
Since the success boundary is reached, we say the treatment succeeds.

@@

\begin{figure}[bp!]
\centering
\subfloat[enrollment distribution]{\includegraphics[width=0.5\textwidth]{snb-first-plot.pdf}}
\hfill
\subfloat[effect of varying $p$]{\includegraphics[width=0.5\textwidth]{mean-and-variance.pdf}}
\caption{
The distribution of a trial that stops after seven patients respond to treatment
or 11 patients do not. (a) shows the distribution when
$p = 0.2$. (b) shows the mean and variance of the distribution varying $p$ 
between zero and one.
}
\label{fig:exp-and-var}
\end{figure}

More generally the distribution of the number of trial enrollments is shown in 
Fig.~\ref{fig:exp-and-var} (a). There is relatively little probability mass
for values of seven through 10 since $p$ is small and the treatment succeeding
quickly is unlikely.
%The probability mass is concentrated at the 
%$11^{\text{th}}$ and $12^{\text{th}}$ step since $p$ 
%is small and for most realizations reach 11 
%non-responders before reaching 2 responders.
Fig.~\ref{fig:exp-and-var} (b) shows the expected value and variance for the
number of trial enrollments varying $p$ between zero and one. When $p$ is
small the probability of an individual response is small, the treatment
is more likely to fail shortly after the $11^{\text{th}}$ enrollment.
When $p$ is large the treatment is more likely to succeed and the number of 
enrollees approaches seven from above. 
Since $p=1$ and $p=0$ represent the deterministic
case the corresponding variance is zero. The ``bump'' around $p=0.25$ results 
from inequality in the endpoints.

@@

In the rest of this work we derive the distribution of the number of 
enrollees needed
to observe either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB). Some of its characteristics are
summarized in Tab.~\ref{tab:snb}.
This paper derives these results
and explores properties of the distribution.
%The next section introduces our notation and basic results
%including the density of the distribution along with a description of
%its relation to other distributions. 
Section 2 derives the distribution function
based on a defined Bernoulli process and gives some basic properties.
Section 3 shows how the distribution is related to other standard
distributions and connects the SNB tail probability to the binomial tail 
probability.
Section 4 derives the moment generating function.
Section 5 derives the compound distribution when $p$ has a beta distribution.

@@

Let $\,b_1, b_2, \ldots \,$ denote a sequence of independent, identically
distributed, Bernoulli random variables with $\mathbb{P}[b_i=1]=p$ and
$\mathbb{P}[b_i = 0] = 1-p$, for
probability parameter $0\leq p \leq 1$. In the clinical trial setting
$\,b_i = 1$ corresponds to a patient responding to treatment.  
Let $s$ and $t$ be positive integers.  Define the SNB random
variable $Y$ as the smallest
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either}
$\,s\,$ responders {\em or} $\,t\,$ non-responders. That is, the SNB 
distribution of $Y$ is the smallest integer such that either
$\sum_i^Y b_i = s$ or $\sum_i^Y 1-b_i = t$.

The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}.
\end{equation*}
The probability mass function is
\begin{equation} \label{eqn:pmf}
\mathbb{P} [Y=k] = S(k, p, s) \ I_{\{s \leq k\}} + 
  S(k, 1-p, t) \ I_{\{ t \leq k \}}
\end{equation}
where $I_{\{f\}}$ is the {\em indicator function}, taking the value 
of one if $f$ is true and zero otherwise, and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is the negative binomial probability mass.

To prove (\ref{eqn:pmf}), consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
At each step a patient's outcome is measured. In Fig.~\ref{fig:kane_viz} 
we consider a graphical illustration of the plot $X_k$ against
$k$. If the outcome of the $k$th patient responds to treatment then the process 
advances diagonally in the positive horizontal and vertical direction. 
If the $k$th patient does not respond
then the sample path advances in the positive horizontal direction only. The
process continues until either $X_k = s$ or $X_k = k-t$.

@@

The probability a given realization of $\mathbf{X}$ reaches $s$ at
the $k$th outcome is the probability that, at time $k-1$, there are $s-1$
successful outcomes and $k-s$ unsuccessful outcomes multiplied by
the probability of a final success at time $k$. This expression is given
in (\ref{eqn:N}). 
Similarly, the probability a given realization reaches $k-t$
is the probability that, at outcome $k-1$, there are $k-t$ successful outcomes
and $t-1$ unsuccessful outcomes multiplied by the probability of a final
unsuccessful outcome at time $k$.

@@

Next, we consider an interim analysis of a clinical trial after $s'$ 
patients respond to treatment 
and $t'$ do not for $s' < s$ and $t' < t$.
\begin{corollary} \label{conditional_distribution}
The number of subsequent enrollments needed 
to reach either endpoint behaves as SNB($p$, $s-s'$, $t-t'$).
\end{corollary}
Having observed $s'$ responders and $t'$ non-responders, there are $s-s'$ 
more responders needed to reach the success endpoint and $t-t'$ more 
non-responders needed to reach the failure endpoint.

\section{Connections and Approximations to Other Distributions}

\begin{figure}[p!]
\begin{center}
\includegraphics[width=\textwidth]{shapes.pdf}
\end{center}
\caption{Different shapes of the SNB distribution with parameters 
($p$, $s$, $t$), as given. Black indicates mass contributed by reaching
$s$ responders before $t$ non-responders. Grey indicates
mass contributed by reaching $t$ non-responders first. \label{shapes.fig}}
\end{figure}

The SNB is a generalization of the negative 
binomial distribution. When $t$ is large $Y-s$ has a 
negative binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j\ |\ t \text{ is large\ }]        \label{nb1.eq}          
  \rightarrow = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is small. As a result, with proper parameter choice, the SNB
can mimic other probability distributions in a manner similar to 
those described in \cite{Peizer1968} and \cite{Best1974}. Examples are
shown in in Fig.~\ref{shapes.fig}.

The SNB is the 
number of enrollment outcomes to observe {\em either} $s$ responders {\em or} 
$t$ non-responders.
The SNB generalizes both the minimum (riff-shuffle) and maximum negative
binomial distributions up to a translation of the support.
For the special case of $\,s=t,$ the distribution of $\,Y\,$ is exactly the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1970}.
The maximum negative binomial \cite{Zhang2000,Zelterman2005} is the smallest 
number of outcomes necessary to 
observe at least $s$ responders {\em and} $s$ non-responders and is equivalent
to a translated version of the riff-shuffle.

@@

%There is a close connection between the tail probabilities of the SNB and the 
%binomial distributions.
%The probability of reaching the success endpoint in an 
%SNB($p$, $s$, $t$) random variable is 
%equal to the probability of at least $s$ successes in a binomial distributed 
%random variable with size $s+t-1$ and success probability $p$. 
%Likewise, the probability of reaching the failure endpoint is equal
%to the probability of at most $s-1$ successes in a binomial distribution. 
There is an equivalence between the probability of reaching an endpoint 
in the SNB model and the tail probability of the binomial distribution.
That is, the probability that the number of responders is at least $s$ in the 
binomial model is the same that the treatment succeeds (reaches $s$) in the SNB 
model.
\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed binomial with size $n=s+t-1$ and response probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [\text{treatment succeeds}].
\end{equation}
\end{prop}
\begin{proof}
The binomial tail probability is
\begin{equation*}
\mathbb{P}[B \geq s] = 1 - \mathcal{I}_{1-p}.
\end{equation*}
%\begin{align*}
%\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \mathcal{I}_{1-p}.
%\end{align*}
The corresponding SNB probability is
\begin{equation*}
\mathbb{P} [\text{treatment succeeds}] 
  = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{equation*}
Let $i=k-s$. Using the fact that
\begin{equation*}
{i+s-1 \choose s-1} = {i+s-1 \choose i}
\end{equation*}
the last summation can be rewritten as
\begin{align}
\mathbb{P} [\text{treatment succeeds}] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align}
completing the proof.
\end{proof}

To illustrate this result, let us return to our initial example
where $s=7$, $t=11$, and $p=0.2$.  The probability masses in
Fig.~\ref{fig:snb_bin_compare} represented in 
black are equal in panels (a) and (b) as are the masses in grey.
The probability that $s$
responders are reached in the SNB process is the same as the binomial 
probability of at least seven responders. Likewise, the probability that $t$ 
non-responders are reached in the SNB process is the same as the binomial
probability of zero through six responders.

\begin{figure}[t!]
\centering
\subfloat[SNB distribution]{\includegraphics[width=0.5\textwidth]{snb_density.pdf}}
\hfill
\subfloat[binomial distribution]{\includegraphics[width=0.5\textwidth]{bin_density.pdf}}
\caption{
SNB(0.2, 7, 11) with mass contributed from 
$s$ responders (black) or $t$ non-responders (grey) along with 
Binomial(0.2, 17) with at least 2 responders (black) or fewer (grey).
}
\label{fig:snb_bin_compare}
\end{figure}

@@

The SNB's ability to approximate the geometric, normal, gamma, and poisson
distributions follow from it generalizing the negative binomial distribution. 
To recover the MGF of the negative binomial consider the case where
$t \rightarrow \infty$ in (\ref{eqn:mgf}). The regularized incomplete
beta function in the first term goes to one and zero in the second term.
We are left with the MGF of the negative binomial distribution.

When $s=1$ (and $t$ is still large) the SNB's MGF is the same as that of the
geometric distribution. The negative
binomial can therefore be seen as a sum of i.i.d. geometric distributions.
For an appropriately large number of samples the central limit theorem
yields a normal approximation.

Drawing connections to the gamma and poisson distributions are more complicated.
However a connection to the gamma distribution well-studied problem in the 
literature (see \cite{Ord1968,Guenther1972,Best1974} for examples). 
A connection to the poisson appears in \cite{Anscombe1950} where
it is shown that if the mean of a poisson is proportional
to a $\chi^2_{2k}$ distribution then the negative binomial is obtained. 
Both of these approximations work by equating cumulants and then showing 
that differences between betwen the cumulant generating
functions converge to zero.

@@

Let us consider the SNB distribution where $p$ has a beta distribution.
\begin{prop} \label{prop:bayesian}
The compound distribution of the SNB
distribution when $p$ is distributed Beta($\alpha$, $\beta$) is
\begin{align} \label{eqn:posterior}
\mathbb{P} [Y = k | s, t, \alpha, \beta ] &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose t-1} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}.
\end{align}
\end{prop}
\begin{proof}
For notational simplicity, assume that $k$ is in the range
$\min(s,t) \leq k \leq s+t-1$. When 
this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose t-1} p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose t-1} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp.
\end{align*}
The result in (\ref{eqn:posterior}) follows by the definition of the 
beta function.
\end{proof}

To better understand the compound distribution, let us once again return to 
our example where $s=7$, $t=11$. However, we assume $p$ is a beta distribution
with shape parameters $\alpha = 0.2 c$ and 
$\beta = 0.8 c$ for any $c > 0$.
This parameterization allows us to examine the relationship between 
uncertainty in $p$ and the SNB.

\begin{figure}[h!]
\centering
\subfloat[mean]{\includegraphics[width=0.5\textwidth]{bayesian-sample-expectation.pdf}}
\hfill
\subfloat[variance]{\includegraphics[width=0.5\textwidth]{bayesian-sample-variance.pdf}}
\caption{
The mean and variance of the compound SNB with $s=7$, $t=11$
with $\mathbb{E}\ p = 0.2$ varying the uncertainty 
in the distribution of $p$. A larger value of $c$
corresponds to a beta with smaller variance.
}
\label{fig:bayesian-sample-size}
\end{figure}

Fig.~\ref{fig:bayesian-sample-size} shows how the mean
and variance of the compound distribution is related to the SNB's 
mean and variance. As $c$ increases the 
compound distribution sample statistics converge to corresponding 
statistics for the SNB with $p = 0.2$. 
%It may be noted that the compound 
%distribution's mean
%approaches the deterministic limit slowly. This
%is because when $c$ is small, the distribution of $p$ is heavily skewed. 
%As $c$ increases this effect is minimized.
When the distribution of $p$ is centered close to zero or one the mean 
estimate approach the deterministic limit relatively slowly in the uncertainty
parameter $c$.  Because of this, we note that for estimation it may be 
preferable to select shape parameters based on the median or mode of the 
beta distribution since they are less sensitive to distributional uncertainty.

@@

%A second avenue for methodological innovation comes directly from
%Corollary \ref{conditional_distribution} and Proposition \ref{prop:bayesian}.
%Our knowledge of the response probability $p$ comes from published reports
%yielding a prior distribution.
%While a trial is being conducted the response probabilities are 
%updated as new results are received. The number of individual successes
%and failure are used to learn the shape characteristics of the corresponding
%beta distribution and resulting posterior
%distribution. An estimate of the probability of reaching the success (and
%failure) endpoints can be provided at each step of the trial.
